Energy stable discontinuous Galerkin formulations using discrete differential operators

### 2017-05-09 docs/splitform/main.tex
updates to elasticity notes (sisc rejected reviewer comments)

\usepackage{fullpage}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem*{remark}{Remark}
%\usepackage{thmtools}
%\declaretheorem[style=definition,qed=$\blacksquare$,numberwithin=chapter]{definition}

@@

\date{}
\author{Jesse Chan}
\title{Energy stable discontinuous Galerkin formulations and discrete differential operators}

@@

We assume that the domain $\Omega$ is decomposed into non-overlapping elements $D^k$.  We define the mesh $\Omega_h = \cup D^k$ and the corresponding global approximation space $V_h(\Omega_h) = \bigoplus V_h\LRp{D^k}$, where $V_h\LRp{D^k}$ is the approximation space over a single patch.  Furthermore, we introduce the jump and average of discontinuous functions across element interfaces.  Let $D^{k,+}$ denote the neighboring element of across a face $f$ of $D^k$, and let $u^+,u^-$ denote the values of $u$ on $D^{k,+}$ and $D^k$, respectively.  The jump of $u$ across $f$ is then defined as
\[
\jump{u} = u^+ - u^-, \qquad \avg{u} = \frac{u^+ + u^-}{2}.
\]
The jump and average of vector fields are defined component-wise using the jumps and averages of components.

@@

In order to define discrete differential operators, we first introduce the $L^2$ inner product on $V_h\LRp{\Oh}$
\[
\LRp{u,v}_{\Omega} = \sum_{D^k} \int_{D^k} uv \diff{x} = \sum_{D^k} \int_{\widehat{D}} uv J \diff{\widehat{x}}.  
\]
In practice, these integrals are computed using quadrature, such that 
\[
\int_{\widehat{D}}uv = \sum_{i=1}^{N_q} u(\bm{x}_i)v(\bm{x}_i) w_i,
\]
where $N_q$ is the number of quadrature points.  The only assumptions we make upon this quadrature is that it is sufficiently accurate such that 
\begin{enumerate}
\item The quadrature induces an equivalent inner product over the reference element $\widehat{D}$
\item Integration by parts holds (with respect to the reference coordinates $\widehat{x}$).
\end{enumerate}

We also introduce the $L^2$ projection $\Pi_N: L^2\LRp{\Omega}\rightarrow P^N$ such that 
\[
\LRp{\Pi_N u,v}_{\Omega} = \LRp{u,v}_{\Omega}.
\]

\subsection{Discrete derivatives}
We introduce two discrete DG derivatives in this section.  The first is the implicit definition used in \cite{di2011mathematical}

\begin{definition}
The discontinuous Galerkin differentiation operator ${D}^i: L^2\LRp{\Omega}\rightarrow V_h$ is defined implicitly as follows:
\[
\LRp{{D}^i u,v}_{\Oh} = \sum_{D^k} \LRp{\LRp{-u,\pd{v}{\bm{x}_i}}_{\LK} + \LRa{\avg{u},v\bm{n}_i}_{\LdK}}.
\]
\end{definition}

We note that, for $u,v \in V_h$, integration by parts yields an equivalent definition of $D^i$ found in \cite{hesthaven2004high,Warburton20063205}
\[
\LRp{{D}^i u,v}_{\Oh} = \sum_{D^k} \LRp{\LRp{\pd{u}{\bm{x}_i},v}_{\LK} + \frac{1}{2}\LRa{\jump{u},v\bm{n}_i}_{\LdK}}.
\]
However, for $u \not\in V_h$, these two definitions are not equivalent.  This motivates the definition of a \textit{discrete} DG differentiation operator:
\begin{definition}
The discrete discontinuous Galerkin differentiation operator ${D}^i_h: L^2\LRp{\Omega} \rightarrow V_h$ is defined as $D^i_h = D^i \Pi_N$.
\end{definition}

The main property of

@@

When the support of $v$ is limited to a single element, we have
\[
\LRp{\Grad_h\cdot \bm{u},v\bbm{1}_{D^k}} = \LRp{\bm{u}, \Grad\cdot v\bbm{1}_{D^k}}  + \LRa{\avg{\bm{u}}\cdot\bm{n},v}_{\partial D^k}.
\]
and as a result when $v = 1$
\[
\LRp{\Grad_h\cdot \bm{u},\bbm{1}_{D^k}} = \int_{\partial D^k}\avg{\bm{u}}\cdot\bm{n}
\]
\section{Variable advection}

\section{Local conservation}

@@

@@

Can also write the inconsistent term as 
\[
\LRp{-{ \bm{\beta}u},\Grad_h v} %= -\LRp{\Grad_h v,\bm{\beta}u} = \sum_{D^k} \LRp{ v,\Grad\cdot \LRp{\bm{\beta}u}}_{D^k} - \LRa{\avg{v},\bm{\beta}\cdot \bm{n} u}
\]
but this results in the same inconsistent flux term due to the fact that $\Grad_h$ is only defined for test functions in the polynomial space.

@@

For curvilinear coordinates,

@@

\section{Standard entropy stability estimates}

Given a nonlinear conservation law
\begin{align*}
\pd{u}{t} + \pd{f(u)}{x} = 0
\end{align*}
DG formulation is usually
\[
\LRp{\pd{u}{t},v} + \LRp{D_h f,v} = 0.
\]
Taking $v = u$ gives
\[
\frac{1}{2}\nor{u}^2 + \int_{D^k}{u D_h f} = 0.
\]
Continuous entropy stability relies on the introduction of $F,G$ such that
\[
\pd{F}{u} = u\pd{f}{u}, \qquad \pd{G}{u} = f, \qquad F = uf - G.
\]
Then, relying on product and chain rules
\[
\int_{D^k}{u \pd{f}{x}} = \int_{D^k}{\pd{(uf)}{x} - f\pd{u}{x}} = \int_{D^k}{\pd{(uf)}{x} - \pd{G}{u}\pd{u}{x}} =  \int_{D^k}{\pd{\LRp{uf-G}}{x}} = \int_{D^k}{\pd{F}{x}} = \int_{\partial D^k} F.
\]
These boundary terms can be made to cancel with appropriately defined numerical fluxes, resulting in entropy conservative schemes.  The challenge in reproducing this in the discrete case is the lack of a product and chain rule for inexact quadratures.  Jameson deals with this by introducing an integrated flux
\[
G = \frac{1}{2}\int_{-1}^1 f \LRp{ \avg{u} + \theta \jump{u}} \diff \theta.
\]
for which finite volume schemes satisfy (in a rough sense)
\[
\jump{G} = \avg{G} \jump{u}.
\]
Fischer and Carpenter (also Gassner and co-workers) use a similar idea but combine a symmetric two-point flux approximation with properties of SBP matrices to get 
\[
2\LRp{W,DF} = \LRp{W,DF} - \LRp{DW,F} + \LRa{W,F} =  \sum_i \sum_j (W_i-W_j) (MD)_{ij} F(U_i,U_j) + \LRa{W,F}.
\]

\section{Entropy splitting}

It's currently unclear how to extend the symmetric two-point flux approximation to quadrature-based DG methods.

>>>>

\usepackage{fullpage}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{lemma}
\newtheorem{lemma}{Lemma}
\newtheorem*{remark}{Remark}
%\usepackage{thmtools}
%\declaretheorem[style=definition,qed=$\blacksquare$,numberwithin=chapter]{definition}

@@

\date{}
\author{Jesse Chan}
\title{Energy stable discontinuous Galerkin formulations using discrete differential operators}

@@

We assume that the domain $\Omega$ is decomposed into non-overlapping elements $D^k$.  We define the mesh $\Omega_h = \cup D^k$ and the corresponding global approximation space $V_h(\Omega_h) = \bigoplus V_h\LRp{D^k}$, where $V_h\LRp{D^k}$ is the approximation space over a single patch.  Furthermore, we introduce the jump and average of discontinuous functions across element interfaces.  Let $D^{k,+}$ denote the neighboring element of across a face $f$ of $D^k$, and let $u^+,u^-$ denote the values of $u$ on $D^{k,+}$ and $D^k$, respectively.  The jump of $u$ across $f$ is then defined as
\[
\jump{u} = u^+ - u^-, \qquad \avg{u} = \frac{u^+ + u^-}{2}.
\]
On faces which coincide with the boundary $\partial \Omega$, the average and jump are defined as
\[
\jump{u} = 0, \qquad \avg{u} = u.
\]
The jump and average of vector fields are defined component-wise using the jumps and averages of components.

We will assume for now that all elements are affine mappings of a reference element $\widehat{D}$.

@@

In order to define discrete differential operators, we first introduce the $L^2$ inner product on $V_h\LRp{\Oh}$
\[
\LRp{u,v}_{\Omega} = \sum_{D^k} \int_{D^k} uv \diff{x} = \sum_{D^k} \int_{\widehat{D}} uv J \diff{\widehat{x}},
\]
and the associated $L^2$ projection $\Pi_N: L^2\LRp{\Omega}\rightarrow P^N$
\[
\LRp{\Pi_N u,v}_{\Omega} = \LRp{u,v}_{\Omega}.  
\]

In practice, these integrals are computed using quadrature, such that 
\begin{equation}
\int_{\widehat{D}}uv = \sum_{i=1}^{N_q} u(\bm{x}_i)v(\bm{x}_i) w_i,
\label{eq:quad}
\end{equation}
where $N_q$ is the number of quadrature points.  The only assumptions we make upon this quadrature is that it is sufficiently accurate such that 
\begin{enumerate}
\item The quadrature induces an $L^2$-equivalent inner product over the reference element $\widehat{D}$.
\item The quadrature is sufficiently accurate such that integration by parts holds with respect to the reference coordinates $\widehat{x}$.
\end{enumerate}

\subsection{Discrete derivatives}
We introduce two discrete DG derivatives in this section.  The first is the implicit definition used in \cite{di2011mathematical}:
\begin{definition}
The discontinuous Galerkin differentiation operator ${D}^i: L^2\LRp{\Omega}\rightarrow V_h$ is defined implicitly as follows:
\[
\LRp{{D}^i u,v}_{\Oh} = \sum_{D^k} \LRp{\LRp{-u,\pd{v}{\bm{x}_i}}_{\LK} + \LRa{\avg{u},v\bm{n}_i}_{\LdK}}, \qquad \forall v \in V_h.
\]
\end{definition}

We note that, for $u,v \in V_h$, (discrete) integration by parts yields an equivalent definition of $D^i$ found in \cite{hesthaven2004high,Warburton20063205}
\[
\LRp{{D}^i u,v}_{\Oh} = \sum_{D^k} \LRp{\LRp{\pd{u}{\bm{x}_i},v}_{\LK} - \frac{1}{2}\LRa{\jump{u},v\bm{n}_i}_{\LdK}}, \qquad \forall v\in V_h.
\]
The main property of the discrete derivatives utilized in this work is a discrete integration-by-parts formula.  
\begin{lemma}
Let $u,v\in V_h$.  Then, 
\begin{align*}
\LRp{D^i u,v} &= \LRp{-u,D^i_h v} + \int_{\partial \Omega} uv \bm{n}_i\\
\LRp{D_h^i u,v} &= \LRp{-u,D^i_h v} + \int_{\partial \Omega} uv \bm{n}_i.
\end{align*}
\label{lemma:ibp}
\end{lemma}
\begin{proof}
Ez
\end{proof}

For $u \not\in V_h$, due to the fact that the quadrature rule (\ref{eq:quad}) may not be exact, and these two definitions are not equivalent and Lemma~\ref{lemma:ibp} does not hold.  This motivates the definition of a \textit{discrete} DG differentiation operator:
\begin{definition}
The discrete discontinuous Galerkin differentiation operator ${D}^i_h: L^2\LRp{\Omega} \rightarrow V_h$ is defined as $D^i_h = D^i \Pi_N$.
\end{definition}

Using these discrete derivatives, we can define discrete gradient and divergence operators
\begin{definition}
Let $u \in V_h$ and $\bm{u}\in \LRp{V_h}^d$.  Then, discrete gradient and divergence operators are defined as follows:
\begin{align*}
\LRp{\Grad^{\rm DG} u}_i &= D^i u, \qquad \Grad^{\rm DG}\cdot \bm{u} = \sum_{i=1}^d D^i \bm{u}_i\\
\LRp{\Grad^{\rm DG}_h u}_i &= D_h^i u, \qquad \Grad^{\rm DG}_h\cdot \bm{u} = \sum_{i=1}^d D_h^i \bm{u}_i.
\end{align*}
\end{definition}

\subsection{Examples}

\begin{itemize}
\item Tensor product elements: inner product computed using GLL nodes, $\Pi_N$ reduces to nodal interpolation
\item General elements: $\Pi_N$ is just quadrature-based projection.  For accuracy, degree $2N$ or higher?  
\item General elements: $\Pi_N$ can be taken to be nodal interpolation.  
\item Pyramids?  
\end{itemize}

@@

When the support of $v$ is limited to a single element, we have
\[
\LRp{\Grad_h\cdot \bm{u},v\bbm{1}_{D^k}} = \LRp{\bm{u}, \Grad\cdot v\bbm{1}_{D^k}}  + \LRa{\avg{\bm{u}}\cdot\bm{n},v}_{\partial D^k}.
\]
and as a result when $v = 1$
\[
\LRp{\Grad_h\cdot \bm{u},\bbm{1}_{D^k}} = \int_{\partial D^k}\avg{\bm{u}}\cdot\bm{n}
\]

\subsection{Local conservation}

@@

\section{Variable advection}

@@

@@

\section{Curvilinear coordinates}

For curvilinear meshes, $J$ and $\bm{G}$ are no longer constant over each element.  However, for isoparametric mappings (and special polynomial interpolations of general curvilinear mappings), metric identities hold \cite{kopriva2006metric} such that $\widehat{\Grad} \cdot \LRp{J\bm{G}} = 0$.  
\[
\LRp{\Grad u,\bm{v}}_{D^k} = \LRp{J\bm{G}\widehat{\Grad} u,\bm{v}}_{\widehat{D}} = \LRp{-u,\widehat{\Grad} \cdot \LRp{J\bm{G}\bm{v}}}  + \LRa{J\bm{G}\widehat{\bm{n}} u,\bm{v}} 
\]
Noting that $J\bm{G}\widehat{\bm{n}} = \bm{n} J^f$ \cite{hesthaven2007nodal}, we recover integration by parts over the physical element.

@@

%\section{Standard entropy stability estimates}
%
%Given a nonlinear conservation law
%\begin{align*}
%\pd{u}{t} + \pd{f(u)}{x} = 0
%\end{align*}
%DG formulation is usually
%\[
%\LRp{\pd{u}{t},v} + \LRp{D_h f,v} = 0.
%\]
%Taking $v = u$ gives
%\[
%\frac{1}{2}\nor{u}^2 + \int_{D^k}{u D_h f} = 0.
%\]
%Continuous entropy stability relies on the introduction of $F,G$ such that
%\[
%\pd{F}{u} = u\pd{f}{u}, \qquad \pd{G}{u} = f, \qquad F = uf - G.
%\]
%Then, relying on product and chain rules
%\[
%\int_{D^k}{u \pd{f}{x}} = \int_{D^k}{\pd{(uf)}{x} - f\pd{u}{x}} = \int_{D^k}{\pd{(uf)}{x} - \pd{G}{u}\pd{u}{x}} =  \int_{D^k}{\pd{\LRp{uf-G}}{x}} = \int_{D^k}{\pd{F}{x}} = \int_{\partial D^k} F.
%\]
%These boundary terms can be made to cancel with appropriately defined numerical fluxes, resulting in entropy conservative schemes.  The challenge in reproducing this in the discrete case is the lack of a product and chain rule for inexact quadratures.  Jameson deals with this by introducing an integrated flux
%\[
%G = \frac{1}{2}\int_{-1}^1 f \LRp{ \avg{u} + \theta \jump{u}} \diff \theta.
%\]
%for which finite volume schemes satisfy (in a rough sense)
%\[
%\jump{G} = \avg{G} \jump{u}.
%\]
%Fischer and Carpenter (also Gassner and co-workers) use a similar idea but combine a symmetric two-point flux approximation with properties of SBP matrices to get 
%\[
%2\LRp{W,DF} = \LRp{W,DF} - \LRp{DW,F} + \LRa{W,F} =  \sum_i \sum_j (W_i-W_j) (MD)_{ij} F(U_i,U_j) + \LRa{W,F}.
%\]
%
%\section{Entropy splitting}
%
%It's currently unclear how to extend the symmetric two-point flux approximation to quadrature-based DG methods.
